Lattice Boltzmann simulations of the dynamics of thin liquid films \\ \large{Gitter Boltzmann Simulationen der Dynamik von d\"unnen Fl\"ussigkeitsfilmen}

### 2024-08-01 chapters/Method.tex
Updates from Overleaf

\chapter{Method}
\label{chapter:method}
\epigraph{\textit{Du bist schön, aber dafür kannst du nichts.
Weder Lesen, noch Schreiben, noch was anderes.}}{Alligatoah}
Based on the theory discussed in the previous chapter it is possible to build a model for numerical simulations.
Numerical simulations have been at the heart of fluid dynamics for a long time.
This is mainly because of two statements.
On the one hand we cannot find an analytical solution to every given flow problem, which is due to the shape and the structure of the Navier-Stokes equations.
In fact there are only a few flows that admit an analytical solution. 
They are often referred to as benchmark problems.
One of those is the famous ''Hagen–Poiseuille flow'', see  Sec.~\ref{sec:theory_shallow_water}, which is an analytical solution for a flow field in a canal and was first derived by Poiseuille and Haagen in the 18th century~\cite{suteraHistoryPoiseuilleLaw}.
There are a few other benchmark problems, especially in the low Reynolds number regime with a small or vanishing Stokes number\footnote{The Stokes number is a dimensionless number that defines the behaviour of particles suspended in a flow and can be calculated with $St = \frac{t_0u_0}{l_0}$, where $t_0$ is a relaxation time of the particles, $u_0$ is a characteristic velocity and $l_0$ is a characteristic length scale.}.
The so called Stokes flows admit time reversal symmetry and many have theoretical solutions.
However, not all flows are laminar and in fact most natural flow admit some turbulent behaviour. 
Understanding flows with high Reynolds numbers still poses a problem, or as R. Feynman once said: ''... the most important unsolved problem of classical physics.''

@@

Discretization of a derivative is a long studied problem~\cite{booleTreatiseCalculusFinite1872, jordanCalculusFiniteDifferences1965}.
The Taylor expansion gives us a tool to approximate the solution to a differential.
In one spatial dimension there are three different methods to compute the derivative, neglecting boundary conditions there is
\begin{equation}\label{eq:forward_dif}
    \partial_x f(x_i) = \frac{f(x_{i+1}) - f(x_i)}{h},
\end{equation}
where $h$ is the spacing between two consecutive points as introduced in Fig.~\ref{fig:finite_difference}. 
Subtracting the value of the function $f(x_{i+1})$ from $f(x_i)$ is called forward difference.
Changing the interval the backward difference is given as
\begin{equation}\label{eq:backward_dif}
    \partial_x f(x_i) = \frac{f(x_i) - f(x_{i-1})}{h}.
\end{equation}
Both the forward as well as the backward differential are first order in their error, beyond first order is the central difference approach
\begin{equation}\label{eq:center_diff}
    \partial_x f(x_i) = \frac{f(x_{i+1}) - f(x_{i-1})}{2h},
\end{equation}
where the value of the derivative is averaged over a larger interval. 
The second derivative is then simply applying the differential operator (forward, backward, central) to  the first derivative. 
In case of central differences this yields
\begin{equation}\label{eq:second_central_diff}
    \partial_x^2 f(x_i) = \frac{f(x_{i+1})-2f(x_i)+f(x_{i-1})}{^2}.
\end{equation}

@@

\subsection{The Boltzmann equation and macroscopic conservation equations}
The dynamics of these distributions $f$ can be formalized and put into an evolution equation which is called the Boltzmann equation
\begin{equation}\label{eq:boltzmann_eq}
    \partial_t f + \vec{\xi}\cdot\vec{\nabla} f + \frac{1}{\rho}\vec{F}\cdot\partial_{\vec{\xi}}f = \Omega(f), 
\end{equation}
where $\vec{\xi}$ is the momentum. 
The first two terms describe the advection of the distribution and the third term measures the effect of a force $\mathbf{F}$ on the momentum~\cite{krugerLatticeBoltzmannMethod2017}.
On the right hand side of the equation is an arbitrarily complex term, the collision operator.
This operator accounts for all collisions the particles perform at any given instance of time.
In principle this operator cannot be computed analytically, because one would need to take into account all possible collision terms from two up to n-point interactions.

@@

The discretization of Eq.~(\ref{eq:boltzmann_eq}) using the lattice Boltzmann method is straightforward and does not require complex meshing.
In fact this is one of the strengths the lattice Boltzmann method inherited from the LGA.
As the name suggests a lattice with lattice constants $\Delta x$ and $\Delta t$ for spatial and temporal discretization is the desired numerical domain. 
The discretized Boltzmann equation without force term reads,
\begin{equation}\label{eq:LBM_discret_noforces}
    f_i(\vec{x}+\vec{c}_i\Delta t, t+ \Delta t) - f_i(\vec{x}, t) = -\frac{\Delta t}{\tau}(f_i(\vec{x}, t) - f^{eq}_i(\vec{x}, t)),
\end{equation}
where the $\vec{c}_{i}$ form a discrete set of $\alpha$ velocities that upon symmetry considerations replace $\vec{\xi}$ of the previous section~\cite{rubinsteinTheoryLatticeBoltzmann2008}.
This equation is also called lattice Boltzmann equation (LBE). 
The perturbative expansion around the equilibrium distribution, Eq.~(\ref{eq:chap-ens-expansion}) does also hold for a discretized velocity space, therefore
\begin{equation}\label{eq:expansion_f}
    f_i = f_i^{eq} + \epsilon f_i^{(1)} + \epsilon^2 f_i^{(2)} + O(\epsilon^3).
\end{equation}
Introducing the non-equilibrium part of the discretized distribution $f_{i}$ as $f_i^{neq} = f_i - f_i^{eq}$. 
From Eqs.~(\ref{eq:BGK_1}-\ref{eq:BGK_2}) we know that the non-equilibrium part has to satisfy, 
\begin{align}\label{eq:non_constraint}
    \sum_i f_i^{neq} = 0, \\
    \sum_i \vec{c}_i f_i^{neq} = \vec{0}, 
\end{align}
where due to discretization the integral becomes a sum.
The moments of the equilibrium functions obey the following relations~\cite{chenLatticeBoltzmannMethod1998} 
\begin{align}
    \Pi^{eq} &= \sum_{i} f_{i}^{eq} = \rho, \label{eq:moments_equilibria_1}\\
    \Pi^{eq}_{\alpha} &= \sum_i c_{i\alpha}f_i^{eq} = \rho u_{\alpha}, \label{eq:moments_equilibria_2}\\
    \Pi^{eq}_{\alpha\beta} &= \sum_{i} c_{i\alpha} c_{i\beta}f_i^{eq} = \rho u_{\alpha} u_{\beta} + \rho c_s^2\delta_{\alpha\beta}, \label{eq:moments_equilibria_3}\\
    \Pi^{eq}_{\alpha\beta\gamma} &= \sum_i c_{i\alpha} c_{i\beta} c_{i\gamma}f_i^{eq} = \rho c_s^2(u_{\alpha}\delta_{\beta\gamma} + u_{\beta}\delta_{\gamma\alpha} + u_{\gamma}\delta_{\alpha\beta}), \label{eq:moments_equilibria_4}
\end{align}
where the Greek indices refer to the spatial dimensions $(x,y,z)$. 
For the zeroth and first moment we also require that
\begin{align}
    \sum_{i} f_{i}^{eq} &= \sum_{i} f_{i} = \rho, \\
    \sum_i c_{i\alpha}f_i^{eq} &= \sum_i c_{i\alpha}f_i = \rho u_{\alpha}.
\end{align}
Clearly Eqs.~(\ref{eq:moments_equilibria_1}-\ref{eq:moments_equilibria_4}) resemble Eqs.~(\ref{eq:boltzmann_moment_1}-\ref{eq:boltzmann_moment_4}) at zeroth order of Eq.~(\ref{eq:expansion_f}).
As a solvability condition for the upcoming derivation we use,
\begin{align}
    \sum_{i} f_{i}^{(n)} = 0, \label{eq:all_order_constraint_1}\\
    \sum_{i} \vec{c}_{i}f_{i}^{(n)} = \vec{0},  \label{eq:all_order_constraint_2}
\end{align}
which basically ensures mass and momentum conservation of the collision operator.

@@

The lattice speeds introduced in Eq.~(\ref{eq:speeds_method}) are already two dimensional and so are the weights $W_i$, see Eq.~(\ref{eq:weightsD2Q9_meth}), therefore the last missing piece are the discretized equilibrium distributions $f_i^{eq}$.
In principle we could adopt a form similar to Eq.~(\ref{eq:eqi_dist2}), however Dellar showed this approach is unstable and in fact suggested to use the equilibrium distributions derived by Salmon~\cite{dellarNonhydrodynamicModesPriori2002, salmonLatticeBoltzmannMethod1999},
\begin{align}
    f^{eq}_0 &= h + w_0h\left(-\frac{15}{8}gh - \frac{3}{2}u^2\right),\\
    f^{eq}_i &= w_i h\left(\frac{3}{2}gh + 3 c_{i\alpha}u_{\alpha} + \frac{9}{2}(c_{i\alpha}u_{\alpha})^2 - \frac{3}{2}u^2\right), \quad i\neq 0 .
\end{align}

>>>>

\chapter{Method}
\label{chapter:method}
\epigraph{\textit{Du bist schön, aber dafür kannst du nichts.
Weder Lesen, noch Schreiben, noch was anderes.}}{Alligatoah}
Based on the theory discussed in the previous chapter it is possible to build a model for numerical simulations.
Numerical simulations have been at the heart of fluid dynamics for a long time.
This is mainly because of two statements.
On the one hand we cannot find an analytical solution to every given flow problem, which is due to the shape and the structure of the Navier-Stokes equation.
In fact there are only a few flows that admit an analytical solution. 
They are often referred to as benchmark problems.
One of those is the famous ''Hagen–Poiseuille flow'', see  Sec.~\ref{sec:theory_shallow_water}, which is an analytical solution for a flow field in a canal and was first derived by Poiseuille and Haagen in the 18th century~\cite{suteraHistoryPoiseuilleLaw}.
There are a few other benchmark problems, especially in the low Reynolds number regime with a small or vanishing Stokes number\footnote{The Stokes number is a dimensionless number that defines the behaviour of particles suspended in a flow and can be calculated with $St = \frac{t_0u_0}{l_0}$, where $t_0$ is a relaxation time of the particles, $u_0$ is a characteristic velocity and $l_0$ is a characteristic length scale.}.
The so called Stokes flows admit time reversal symmetry and many have theoretical solutions.
However, not all flows are laminar and in fact most natural flow admit some turbulent behaviour. 
Understanding flows with high Reynolds numbers still poses a problem, or as R. Feynman once said: ''... the most important unsolved problem of classical physics.''

@@

Discretization of a derivative is a long studied problem~\cite{booleTreatiseCalculusFinite1872, jordanCalculusFiniteDifferences1965}.
The Taylor expansion gives us a tool to approximate the solution to a differential.
In one spatial dimension there are three different methods to compute the derivative, neglecting boundary conditions there is
\begin{equation}\label{eq:forward_dif}
    \partial_x f(x_i) = \frac{f(x_{i+1}) - f(x_i)}{h},
\end{equation}
where $h$ is the spacing between two consecutive points as introduced in Fig.~\ref{fig:finite_difference}. 
Subtracting the value of the function $f(x_{i+1})$ from $f(x_i)$ is called forward difference.
Changing the interval the backward difference is given as
\begin{equation}\label{eq:backward_dif}
    \partial_x f(x_i) = \frac{f(x_i) - f(x_{i-1})}{h}.
\end{equation}
Both the forward as well as the backward differential are first order in their error, beyond first order is the central difference approach
\begin{equation}\label{eq:center_diff}
    \partial_x f(x_i) = \frac{f(x_{i+1}) - f(x_{i-1})}{2h},
\end{equation}
where the value of the derivative is averaged over a larger interval. 
The second derivative is then simply applying the differential operator (forward, backward, central) to  the first derivative. 
In case of central differences this yields
\begin{equation}\label{eq:second_central_diff}
    \partial_x^2 f(x_i) = \frac{f(x_{i+1})-2f(x_i)+f(x_{i-1})}{h^2}.
\end{equation}

@@

\subsection{The Boltzmann equation and macroscopic conservation equations}
The dynamics of these distributions $f$ can be formalized and put into an evolution equation which is called the Boltzmann equation
\begin{equation}\label{eq:boltzmann_eq}
    \partial_t f + \vec{\xi}\cdot\vec{\nabla} f + \frac{1}{\rho}\vec{F}\cdot\partial_{\vec{\xi}}f = \Omega(f), 
\end{equation}
where $\vec{\xi}$ is a velocity. 
The first two terms describe the advection of the distribution and the third term measures the effect of a force $\vec{F}$ on the momentum~\cite{krugerLatticeBoltzmannMethod2017}.
On the right hand side of the equation is an arbitrarily complex term, the collision operator.
This operator accounts for all collisions the particles perform at any given instance of time.
In principle this operator cannot be computed analytically, because one would need to take into account all possible collision terms from two up to n-point interactions.

@@

The discretization of Eq.~(\ref{eq:boltzmann_eq}) using the lattice Boltzmann method is straightforward and does not require complex meshing.
In fact this is one of the strengths the lattice Boltzmann method inherited from the LGA.
As the name suggests a lattice with lattice constants $\Delta x$ and $\Delta t$ for spatial and temporal discretization is the desired numerical domain. 
The discretized Boltzmann equation without force term reads,
\begin{equation}\label{eq:LBM_discret_noforces}
    f_i(\vec{x}+\vec{c}_i\Delta t, t+ \Delta t) - f_i(\vec{x}, t) = -\frac{\Delta t}{\tau}(f_i(\vec{x}, t) - f^{eq}_i(\vec{x}, t)),
\end{equation}
where the $\vec{c}_{i}$ form a discrete set of $i$ velocities that upon symmetry considerations replace $\vec{\xi}$ of the previous section~\cite{rubinsteinTheoryLatticeBoltzmann2008}.
This equation is also called lattice Boltzmann equation (LBE). 
The perturbative expansion around the equilibrium distribution, Eq.~(\ref{eq:chap-ens-expansion}) does also hold for a discretized velocity space, therefore
\begin{equation}\label{eq:expansion_f}
    f_i = f_i^{eq} + \epsilon f_i^{(1)} + \epsilon^2 f_i^{(2)} + O(\epsilon^3).
\end{equation}
Introducing the non-equilibrium part of the discretized distribution $f_{i}$ as $f_i^{neq} = f_i - f_i^{eq}$. 
From Eqs.~(\ref{eq:BGK_1}-\ref{eq:BGK_2}) we know that the non-equilibrium part has to satisfy, 
\begin{align}\label{eq:non_constraint}
    \sum_i f_i^{neq} = 0, \\
    \sum_i \vec{c}_i f_i^{neq} = \vec{0}, 
\end{align}
where due to discretization the integral becomes a sum.
The moments of the equilibrium functions obey the following relations~\cite{chenLatticeBoltzmannMethod1998} 
\begin{align}
    \Pi^{eq} &= \sum_{i} f_{i}^{eq} = \rho, \label{eq:moments_equilibria_1}\\
    \Pi^{eq}_{\alpha} &= \sum_i c_{i\alpha}f_i^{eq} = \rho u_{\alpha}, \label{eq:moments_equilibria_2}\\
    \Pi^{eq}_{\alpha\beta} &= \sum_{i} c_{i\alpha} c_{i\beta}f_i^{eq} = \rho u_{\alpha} u_{\beta} + \rho c_s^2\delta_{\alpha\beta}, \label{eq:moments_equilibria_3}\\
    \Pi^{eq}_{\alpha\beta\gamma} &= \sum_i c_{i\alpha} c_{i\beta} c_{i\gamma}f_i^{eq} = \rho c_s^2(u_{\alpha}\delta_{\beta\gamma} + u_{\beta}\delta_{\gamma\alpha} + u_{\gamma}\delta_{\alpha\beta}), \label{eq:moments_equilibria_4}
\end{align}
where the Greek indices refer to the spatial dimensions $(x,y,z)$. 
For the zeroth and first moment we also require that
\begin{align}
    \sum_{i} f_{i}^{eq} &= \sum_{i} f_{i} = \rho, \\
    \sum_i c_{i\alpha}f_i^{eq} &= \sum_i c_{i\alpha}f_i = \rho u_{\alpha}.
\end{align}
Clearly Eqs.~(\ref{eq:moments_equilibria_1}-\ref{eq:moments_equilibria_4}) resemble Eqs.~(\ref{eq:boltzmann_moment_1}-\ref{eq:boltzmann_moment_4}) at zeroth order of Eq.~(\ref{eq:expansion_f}).
As a solvability condition for the upcoming derivation we use,
\begin{align}
    \sum_{i} f_{i}^{(n)} = 0, \label{eq:all_order_constraint_1}\\
    \sum_{i} \vec{c}_{i}f_{i}^{(n)} = \vec{0},  \label{eq:all_order_constraint_2}
\end{align}
which basically ensures mass and momentum conservation of the collision operator.

@@

The lattice speeds introduced in Eq.~(\ref{eq:speeds_method}) are already two dimensional and so are the weights $w_i$, see Eq.~(\ref{eq:weightsD2Q9_meth}), therefore the last missing piece are the discretized equilibrium distributions $f_i^{eq}$.
In principle we could adopt a form similar to Eq.~(\ref{eq:eqi_dist2}), however Dellar showed this approach is unstable and in fact suggested to use the equilibrium distributions derived by Salmon~\cite{dellarNonhydrodynamicModesPriori2002, salmonLatticeBoltzmannMethod1999},
\begin{align}
    f^{eq}_0 &= h + w_0h\left(-\frac{15}{8}gh - \frac{3}{2}u^2\right),\\
    f^{eq}_i &= w_i h\left(\frac{3}{2}gh + 3 c_{i\alpha}u_{\alpha} + \frac{9}{2}(c_{i\alpha}u_{\alpha})^2 - \frac{3}{2}u^2\right), \quad i\neq 0 .
\end{align}
